---
title: "univariate_collab.rmd"
author: "Lexie DelViscio"
date: "2/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
trees <- read.csv('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/treedata_subset.csv')
trees
```

 1. Carry out an exploratory analysis using the tree dataset. Metadata for the tree study can be found here. Specifically, I would like you to develop and compare models for species cover for a habitat generalist Acer rubrum (Red maple) and a habitat specialist Abies fraseri (Frasier fir). Because this dataset includes both continuous and discrete explanatory variables use the function Anova in the packages car as such
```{r}
library(car)
```
Anova(my_mod, type=3)
Compare the p-values you observe using the function Anova to those generated using summary.

For each species address the following additional questions:
How well does the exploratory model appear to explain cover?
which explanatory variables are the most important?
do model diagnostics indicate any problems with violations of OLS assumptions?... answered later in this document.
are you able to explain variance in one species better than another, why might this be the case?

Prior to addressing the above questions you will want to restructure and subset the data using the # # following R code:
wish to model species cover across all sampled plots
create site x sp matrix for two species 
```{r}
sp_cov = with(trees, tapply(cover, list(plotID, spcode), 
                            function(x) round(mean(x))))
sp_cov = ifelse(is.na(sp_cov), 0, sp_cov)
sp_cov = data.frame(plotID = row.names(sp_cov), sp_cov)
# create environmental matrix
cols_to_select = c('elev', 'tci', 'streamdist', 'disturb', 'beers')
env = aggregate(trees[ , cols_to_select], by = list(trees$plotID), 
                function(x) x[1])
names(env)[1] = 'plotID'
# merge species and environmental matrices
site_dat = merge(sp_cov, env, by='plotID')
# subset species of interest
abies = site_dat[ , c('ABIEFRA', cols_to_select)]
acer  = site_dat[ , c('ACERRUB', cols_to_select)]
names(abies)[1] = 'cover'
names(acer)[1] = 'cover'
```

loading more libraries
```{r}
library(ggplot2)
library(gridExtra)
library(scatterplot3d)
library(MASS)
```

viewing the data
```{r}
abies
acer
names(abies)
names(acer)
```

boxplots for quantitative----
and their quantiles
```{r}
ggplot(data = abies) + 
  geom_boxplot(mapping = aes(x = disturb, y = cover)) +  
  labs(x = 'Disturbance', y = 'Cover', title = 'abies disturbance versus cover') 
ggplot(data = acer) + 
  geom_boxplot(mapping = aes(x = disturb, y = cover)) +  
  labs(x = 'Disturbance', y = 'Cover', title = 'acer disturbance versus cover') 

# abies quantiles 
quantile(abies$cover[abies$disturb == 'VIRGIN'])
quantile(abies$cover[abies$disturb == 'CORPLOG'])
quantile(abies$cover[abies$disturb == 'SETTLE'])
quantile(abies$cover[abies$disturb == 'LT-SEL'])
# acer quantiles 
quantile(acer$cover[acer$disturb == 'VIRGIN'])
quantile(acer$cover[acer$disturb == 'CORPLOG'])
quantile(acer$cover[acer$disturb == 'SETTLE'])
quantile(acer$cover[acer$disturb == 'LT-SEL'])
```

ABIES original plots----
```{r}
plot(cover ~ elev, data = abies, xlab = 'Elevation (m)',
     ylab = 'Cover', main= 'abies elevation versus cover')

plot(cover ~ tci, data = abies, xlab = 'Topographic Coverage Index',
     ylab = 'Cover', main= 'abies tci versus cover')

plot(cover ~ streamdist, data = abies, xlab = 'Stream Distance (m)',
     ylab = 'Cover', main= 'abies stream distance versus cover')

plot(cover ~ beers, data = abies, xlab = 'Beers',
     ylab = 'Cover', main= 'abies beers versus cover')
```

ACER original plots----
```{r}
plot(cover ~ elev, data = acer, xlab = 'Elevation (m)',
     ylab = 'Cover', main= 'acer elevation versus cover')

plot(cover ~ tci, data = acer, xlab = 'Topographic Coverage Index',
     ylab = 'Cover', main= 'acer tci versus cover')

plot(cover ~ streamdist, data = acer, xlab = 'Stream Distance (m)',
     ylab = 'Cover', main= 'acer stream distance versus cover')

plot(cover ~ beers, data = acer, xlab = 'Beers',
     ylab = 'Cover', main= 'acer beers versus cover')
```

Intercept only models for Abies and Acer, cover as y variable----
```{r}
null_mod_abies = lm(cover ~ 1, data = abies)
null_mod_acer = lm(cover ~ 1, data = acer)     
null_mod_abies
null_mod_acer

mean(abies$cover)
mean(acer$cover)

plot(cover ~ 1, data = abies)
abline(null_mod_abies, lwd = 2)
abline(h = mean(abies$cover), col = 'red', lty = 2, lwd = 2)

plot(cover ~ 1, data = acer)
abline(null_mod_acer, lwd = 2)
abline(h = mean(acer$cover), col = 'red', lty = 2, lwd = 2)
```

create single variable main effect models for each species and variable----
```{r}
# abies
elev_mod_abies = lm(cover ~ elev, data = abies)
tci_mod_abies = lm(cover ~ tci, data = abies)
streamdist_mod_abies = lm(cover ~ streamdist, data = abies)
disturb_mod_abies = lm(cover ~ disturb, data = abies)
beers_mod_abies = lm(cover ~ beers, data = abies)
# acer 
elev_mod_acer = lm(cover ~ elev, data = acer)
tci_mod_acer = lm(cover ~ tci, data = acer)
streamdist_mod_acer = lm(cover ~ streamdist, data = acer)
disturb_mod_acer = lm(cover ~ disturb, data = acer)
beers_mod_acer = lm(cover ~ beers, data = acer)
```

abies single model main effect summaries----
```{r}
summary(elev_mod_abies)
Anova(elev_mod_abies, type= 3)
summary(tci_mod_abies)
summary(streamdist_mod_abies)
summary(disturb_mod_abies)
summary(beers_mod_abies)
# significant p-values = elev, disturb (b/c of VIRGIN), streamdist
```

acer single model main effect summaries----
```{r}
summary(elev_mod_acer)
summary(tci_mod_acer)
summary(streamdist_mod_acer)
summary(disturb_mod_acer)
summary(beers_mod_acer)
# significant p-values = all EXCEPT tci
# beers has a significant effect on acers, but not abies in these models
```

all main effects for both----
```{r}
all_mod_abies = lm(cover ~ elev + tci + streamdist + disturb + beers, data = abies) 
all_mod_acer = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer)

summary(all_mod_abies)
summary(all_mod_acer)
```

identifying outlers and modifying acer because of them----
This code was used to identify the outliers... identify(acer$cover ~ acer$tci, n = 2)
Remove tci outliers
```{r}
plot(cover ~ tci, data = acer, xlab = 'Topographic Coverage Index',
     ylab = 'Cover', main= 'acer tci versus cover')
acer_subset = acer[-c(121, 318), ]
plot(cover ~ tci, data= acer_subset, main="acer subset tci versus cover", xlab='tci', ylab= 'cover')

dim(acer)
dim(acer_subset)
```

plots redone here for convenience... can take out in final probably----
```{r}

plot(cover ~ streamdist, data = acer, xlab = 'Stream Distance (m)',
     ylab = 'Cover', main= 'acer stream distance versus cover')
acer_subset = acer[-c(121, 318, 297,56,187), ]
dim(acer_subset)

plot(cover ~ streamdist, data = acer_subset, xlab = 'Stream Distance (m)',
     ylab = 'Cover', main= 'acer stream distance versus cover subset')
```

new single variable models with the acersubset----
```{r}
elev_mod_acersubset = lm(cover ~ elev, data = acer_subset)
tci_mod_acersubset = lm(cover ~ tci, data = acer_subset)
streamdist_mod_acersubset = lm(cover ~ streamdist, data = acer_subset)
disturb_mod_acersubset = lm(cover ~ disturb, data = acer_subset)
beers_mod_acersubset = lm(cover ~ beers, data = acer_subset)
```

# comparing the original summaries to the subset without outliers----
```{r}
# elevation
summary(elev_mod_acer)
summary(elev_mod_acersubset)
# topographic coverage index
summary(tci_mod_acer)
summary(tci_mod_acersubset)
# stream distance
summary(streamdist_mod_acer)
summary(streamdist_mod_acersubset)
# disturbance
summary(disturb_mod_acer)
summary(disturb_mod_acersubset)
#beers index
summary(beers_mod_acer)
summary(beers_mod_acersubset)
```

comparing the original main model with the subset main model for acer----
```{r}
all_mod_acersubset = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer_subset)
summary(all_mod_acersubset)
summary(all_mod_acer)

# removing the outliers did slightly improve the R-squared and adjusted R- squared, it also greatly increased the significance of tci( lowered the p value), so the subset without the outliers will be used from now on
```

# model with elev, tci, and beers----
etb_mod_acersubset = lm(cover ~ elev + tci + beers , data = acer_subset)
summary(etb_mod_acersubset)

etb_interaction_mod_acersubset = lm(cover ~ elev + tci + beers + tci * elev + elev * beers + beers * tci + beers * tci * elev , data = acer_subset)
summary(etb_interaction_mod_acersubset)

# interaction effect with elev:tci and elev:beers
AIC(etb_interaction_mod_acersubset)
#[1] 3413.76
AIC(etb_mod_acersubset)
#[1] 3419.997
## the above model just proved my suspicions on how important elevation was to the model, it will not be used over other models however because the r squared was lower

#adding elevation interaction to main model-------------
elev_interaction_mod_acersubset = lm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers + elev * streamdist + elev * disturb, data = acer_subset)
summary(elev_interaction_mod_acersubset)

elev_interaction_mod_acersubset = lm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers + elev * streamdist + elev * disturb, data = acer_subset)

# adding interaction to whole model----
full_mod_acersubset = update(all_mod_acersubset, ~ . + elev * disturb * tci * streamdist * beers)
all_mod_acersubset = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer_subset)

summary(all_mod_acersubset)
summary(elev_interaction_mod_acersubset)
summary(full_mod_acersubset) 
# this makes it absurdly long, but it does increase the r-squared...


anova(all_mod_acersubset, full_mod_acersubset)

AIC(full_mod_acersubset)
AIC(all_mod_acersubset)
AIC(elev_interaction_mod_acersubset)
# ele

# remove abies outlier in tci and elevation  
# identify(abies$cover ~ abies$tci, n=2)

abies_subset = abies[-c(121,33,109,56,53), ]
dim(abies_subset)

# comparing the main models of original and subset
all_mod_abiessubset = lm(cover ~ elev + tci + streamdist + disturb + beers, data = abies_subset)
summary(all_mod_abiessubset)
summary(all_mod_abies)


# abies with interaction, full and elevation----
full_mod_abiessubset = update(all_mod_abiessubset, ~ . + elev * disturb * tci * streamdist * beers)
elev_interaction_mod_abiessubset = lm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers + elev * streamdist + elev * disturb, data = abies_subset)

summary(all_mod_abiessubset)
summary(elev_interaction_mod_abiessubset)
summary(full_mod_abiessubset) 

# interaction had an even larger effect on the abies distributions


# Poisson calculation start----
# assume all poisson calculations are done using the subsets with outliers removed
summary(elev_interaction_mod_abiessubset)
summary(elev_interaction_mod_acersubset)

# poisson individual variables acer
elev_mod_acer_poi = glm(cover ~ elev, data = acer_subset, family = "poisson")
tci_mod_acer_poi = glm(cover ~ tci, data = acer_subset, family = "poisson")
streamdist_mod_acer_poi = glm(cover ~ streamdist, data = acer_subset, family = "poisson")
disturb_mod_acer_poi = glm(cover ~ disturb, data = acer_subset, family = "poisson")
beers_mod_acer_poi = glm(cover ~ beers, data = acer_subset, family = "poisson")

# poisson individual variables abies 
elev_mod_abies_poi = glm(cover ~ elev, data = abies_subset, family = "poisson")
tci_mod_abies_poi = glm(cover ~ tci, data = abies_subset, family = "poisson")
streamdist_mod_abies_poi = glm(cover ~ streamdist, data = abies_subset, family = "poisson")
disturb_mod_abies_poi = glm(cover ~ disturb, data = abies_subset, family = "poisson")
beers_mod_abies_poi = glm(cover ~ beers, data = abies_subset, family = "poisson")

# summaries of individual poisson models
# abies
summary(elev_mod_abies_poi)
summary(tci_mod_abies_poi)
summary(streamdist_mod_abies_poi)
summary(disturb_mod_abies_poi)
summary(beers_mod_abies_poi)

# acer
summary(elev_mod_acer_poi)
summary(tci_mod_acer_poi)
summary(streamdist_mod_acer_poi)
summary(disturb_mod_acer_poi)
summary(beers_mod_acer_poi)


# poisson with elevation interaction only----
abies_poi = glm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers + elev * streamdist + elev * disturb, data = abies_subset, 
                family='poisson')
acer_poi = glm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers + elev * streamdist + elev * disturb, data = acer_subset, 
               family='poisson')

summary(abies_poi)
summary(acer_poi)

# pseudo r ^2
pseudo_r2 = function(glm_mod) {
  1 -  glm_mod$deviance / glm_mod$null.deviance
}
pseudo_r2(abies_poi)
pseudo_r2(acer_poi)

# comparing poisson with equivalent OLS----
summary(abies_poi)
summary(elev_interaction_mod_abiessubset)
pseudo_r2(abies_poi)

summary(acer_poi)
summary(elev_interaction_mod_acersubset)
pseudo_r2(acer_poi)

# comparing with anova
Anova(abies_poi, type=3)
Anova(elev_interaction_mod_abiessubset, type=3)
Anova(acer_poi, type=3)
Anova(elev_interaction_mod_acersubset, type=3)

# poisson might be better at estimating abies, but not acer based on the pseudo
# r squared values

reduced <- stepAIC(full_mod_abiessubset)

# model diagnostics with plot of model
# observed vs predicted plot lines 
# change order of levels to choose which is the intercept
# use contrast to visualize level code

# checking plots of the elevation interaction model and removing more outliers---
par(mfrow = c(2,2))
plot(elev_interaction_mod_abiessubset)
# removed a few more outliers
# take out 56, 109,53  from abies subset

plot(elev_interaction_mod_acersubset)
# take out 56 , 187 from acer

# semifinal models to use:----
# elev_interaction_mod_acersubset, elev_interaction_mod_abiessubset
# abies_poi (only includes elevation interaction)
# acer_poi (only includes elevation interaction)

# use stepAIC----
# to check the model and decide if any variables can be removed----
stepAIC(elev_interaction_mod_abiessubset)
# shows that it may make sense to remove elev:beers from the abies model
stepAIC(elev_interaction_mod_acersubset)
# shows that elevation and stream distance interaction may not be neccesarry

# retry with those taken out----
#OLS
abies_test_mod = lm(cover ~ elev + tci + streamdist + disturb + elev * tci + elev * streamdist + elev * disturb, data = abies_subset)
acer_test_mod = lm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers +  elev * disturb, data = acer_subset)
summary(abies_test_mod)
summary(acer_test_mod)
#POISSON
abies_test_poi = glm(cover ~ elev + tci + streamdist + disturb + elev * tci + elev * streamdist + elev * disturb, data = abies_subset, family = "poisson")
acer_test_poi = glm(cover ~ elev + tci + streamdist + disturb + beers + elev * tci + elev * beers +  elev * disturb, data = acer_subset, family = "poisson")
summary(abies_test_poi)
summary(acer_test_poi)
summary(elev_interaction_mod_abiessubset)

# the adjusted r squares show this did help
# after stepAIC abies
AIC(abies_test_mod)
AIC(abies_test_poi)
# original abies
AIC(elev_interaction_mod_abiessubset)
AIC(abies_poi)
# after stepAIC acer
AIC(acer_test_mod)
AIC(acer_test_poi)
# original acer
AIC(elev_interaction_mod_acersubset)
AIC(acer_poi)

# the AIC's also show this was helpful, so we can rename the test models as our
# final models to present... keep in mind these final models are all using the subsetted data without outliers
# the name was just shortened for easier access and readability. 

# final models----
final_abies_model <- abies_test_mod
final_acer_model <- acer_test_mod
final_acer_poisson <- acer_test_poi
final_abies_poisson <- abies_test_poi

# 2) and----
# 1) part: "do model diagnostics indicate any problems with violations of OLS assumptions?"----
# OLS versus poisson for each species
# new model to prove this point with the abies species.
abies_elev_poi = glm(cover ~ elev, data= abies_subset, family="poisson")
# also using elev_interaction_mod_abiessubset = lm(cover ~ elev, data= abies_subset)

# When comparing these plots it confirms suspicion that the Poisson distribution is a better fit
# for the abies species, and this is most likely due to the highly skewed elevation data. Since it can
# be seen in the original (cover ~ elevation) plot for abies, the data is highly concentrated around 0 
# coverage until an elevation of 1500, where the trend turns visually turns into a positive correlation between 
# elevation and abies cover. 

par(mfrow=c(1,1))
plot(cover ~ elev, data= abies_subset)

# This can be explained by the fact that the abies species are fraisir firs and they 
# are highly selective on their habitat condition, they must be further north, or high in elevation, in order to 
# grow. So, when the model lm(cover ~ elev, data= abies[_subset]), or any other model including elevation as a 
# variable describing cover will not be taking into account this highly skewed data as it is working under a normal 
# distribution. This can be seen with the hotpink link on the graph below. This is the residual line for cover ~ elev
# for the abies species under OLS conditions. 

plot(cover ~ elev, data= abies_subset)
abline(elev_interaction_mod_abiessubset, col="hot pink")

# The line trends downwards, which we know to be false as the data trend
# actually increases as elevation increases, after a specific point. However, it can be seen through the blue line on the 
# graph that when the model glm(cover ~ elev, data= abies[_subset], family="poisson") is plotted the regression line is 
# much more in line with what we would expect to predict the data. 

plot(cover ~ elev, data= abies_subset)
abline(elev_interaction_mod_abiessubset, col="hot pink")
abline(abies_elev_poi, col= "light blue")

# The intercept begins around the time the data points begin to increase, and it continues with an upward trend until 
# the end of the graph, just like the data points.

# However, for the acer species,
# when comparing these plots it confirms suspicion that the OLS distribution is a better fit
# for the acer species, the residuals vs leverage graph and scale- location are much more
# better distributed along the line, and more evenly distributed in space during the OLS regression
# rather than the poisson regression. 

par(mfrow= c(2,2))
plot(final_acer_model)
plot(final_acer_poisson)

# more proof that OLS is better for acer, and poisson is better for abies
pseudo_r2(final_acer_poisson)
pseudo_r2(final_abies_poisson)

# Compare these to the output adjusted r squared from normal 
summary(final_acer_model)
summary(final_abies_model)
# this further proves that OLS is better fitted for the Acer species as the pseudo r^2 value
# is less than the normal r^2 value, whereas the pseudo R^2 value for abies is higher than the OLS value

# So, it can be concluded that the best models to represent the cover for species Acer and Abies are:
# the OLS regression of Acer (final_acer_model) and the poisson regression of Abies (final_abies_poisson),
# as the skewed data in the abies species makes it less accurate when using OLS regression.

#' 4. (optional) Examine the behavior of the function stepAIC() using the exploratory 
#' models developed above. This is a very simple and not very robust machine learning 
#' stepwise algorithm that uses AIC to select a best model. By default it does a backward selection routine.

# repeat of the stepAIC used in the model selection above to answer question four.
stepAIC(elev_interaction_mod_acersubset)

# using the output from the stepAIC(elev_interaction_mod_acersubset) it can be used to describe the functionality of stepAIC().
# when used the stepwise function takes a backward approach by first looking at all of the variables included in a model taken in the function.
# By running through the function the first time it is identified that the overall AIC is 1310.9, however after the second output when the function
# is modified it is identified to have an AIC of 1309.36. It can be seen that this is equivalent to the AIC Value corresponding to the elev:streamdist 
# ,it can be concluded that the stepAIC function identifies the AIC value of the model IF the variable it is describing is dropped. From this output it 
# can be further concluded that these are the variables which can be removed to to improve the model in a simple way.So, using the Acer subset, the variables 
# elev:streamdist is the onlt variables that should be considered to be removed. However, in the stepAIC(elev_interaction_mod_abiessubset), shown below, it can be 
# determined that removing elev:beers and beers will both improve the proposed model. 

stepAIC(elev_interaction_mod_abiessubset)
